---
title: "Daily Weather Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Daily Weather Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GHCNr)
library(terra)  # for handling countries geometries
```

# Select GHCNd stations

The station inventory file of GHCNd is stored at <https://www.ncei.noaa.gov/products/land-based-station/global-historical-climatology-network-daily>.
The function `stations()` can read from this source or from a local file, specified with `filename`.
The inventory can also be downloaded to a file using `download_inventory()`.

```{r download-inventory, eval=FALSE}
inventory_file <- download_inventory("~/Downloads/ghcn-inventory.txt")
stations <- stations(inventory_file, variables = "TMAX")
```

```{r read-inventory}
stations <- stations(variables = "TMAX")
```

By specifying `variables = "TMAX"` only the stations that recorded that variable are kept.
Available variables implemented at the moment are precipitation ("PRCP"), minimum temperature ("TMIN"), and maximum temperature ("TMAX").

Stations spanning a time range can be filtered easily.
```{r filter-by-year}
stations <- stations[stations$startYear <= 1990, ]
stations <- stations[stations$endYear >= 2000, ]
stations
```

Spatial filters can also be easily applied.
Spatial boundaries of countries can be downloaded from <https://www.geoboundaries.org/> using the `get_countr(couuntry_code = ...)` function, where `country_code` is the ISO3 code.
```{r get-country}
italy <- get_country("ITA")
```

`get_countries()` can take several ISO3 codes to return a geometry of multiple countries.

```{r spatial-filter}
stations <- filter_stations(stations, italy)
plot(italy)
points(stations[, c("longitude", "latitude")], pch = 20, col = "dodgerblue")
```

# Download daily timeseries

Daily timeseries for a station can be downloaded using the `daily()` function.
In addition to the station ID, `daily()` needs start and end dates of the timeseries.
These should be provided as strings with the format "YYYY-mm-dd", e.g., "1990-01-01".

```{r daily}
daily_ts <- daily(
  stations$station[1],
  paste("1990", "01", "01", sep = "-"),  # shorten a bit as example
  paste(stations$endYear[1], "12", "21", sep = "-"),
  variables = "tmin"
)
daily_ts
```

Multiple stations can also be downloaded at once.
Too many stations will cause the API to fail.
```{r multidaily}
daily_ts <- daily(
  stations$station,
  paste("1990", "01", "01", sep = "-"),  # shorten a bit as example
  paste("1991", "12", "31", sep = "-"),  # shorten a bit as example
  variables = "tmin"
)
daily_ts
plot(daily_ts, "tmin")
```

Note the outliers, which are due to flagged records (see below).

Implmented variables are "tmin", "tmax", and "prcp".
`daily()` returns a table with the value of the variable chosen and associated flags.

## Remove flagged records

Flagged records can be removed using `remove_flagged()`.
In `remove_flagged()` the argument `strict` (dafault = `FALSE`) specifies which flags to include.
Flags always removed are:
```{r flags, echo=FALSE}
as.list(GHCNr:::.flags(strict = FALSE))
```
Whereas setting `strict = TRUE` will cause additional flags to be removed.
```{r flags-strict, echo=FALSE}
setdiff(GHCNr:::.flags(strict = TRUE), GHCNr:::.flags(strict = FALSE))
```

This will also remove the "*_flag=" column.
```{r remove-flagged}
daily_ts <- remove_flagged(daily_ts)
plot(daily_ts, "tmin")
```

# Temporal coverage
Coverage of the timeseries can be calculated using `coverage()`.
```{r daily-coverage}
station_coverage <- coverage(daily_ts)
```

The output is a table with coverage by month and year (`monthly_coverage`), by year (`annual_coverage`), and for the whole time period (`year`).
`annual_coverage` is constant within the same year and `year` is always a constant.
This table is useful to inspect stations that may have problematic timeseries, such as 
```{r low-coverage}
station_coverage[station_coverage$annual_coverage < 0.5, ]
```

In this case, there are several stations that have low coverage and that will be removed.
```{r coverages, fig.width=6, fig.height=4}
with(
  unique(station_coverage[station_coverage$year == 1990, c("station", "year", "annual_coverage")]),
  barplot(annual_coverage, names.arg = station, las = 2, main = "1990")
)
with(
  unique(station_coverage[station_coverage$year == 1991, c("station", "year", "annual_coverage")]),
  barplot(annual_coverage, names.arg = station, las = 2, main = "1991")
)
```

Note that missing years are not shown and that `year_coverage` only calculates the number of years that have been covered, not their internal coverage.

```{r plot-timeseries, echo=FALSE, fig.width=6, fig.height = 4}
palette <- hcl.colors(length(unique(daily_ts$station)), "Roma")
names(palette) <- unique(daily_ts$station)
plot(
  NULL, xlab = "", ylab = "Tmin",
  xlim = range(daily_ts$date),
  ylim = range(daily_ts$tmin, na.rm = TRUE),
  axes = FALSE
)
axis(2, seq(-20, 20, by = 10))
axis(
  1,
  seq(min(daily_ts$date), max(daily_ts$date), length.out = 2),
  labels = c("1990", "1991")
)
for (s in unique(daily_ts$station)) {
  with(
    daily_ts[daily_ts$station == s, ],
    lines(date, tmin, col = palette[s])
  )
}
legend(7300, 29, fill = palette, legend = names(palette), bg = adjustcolor("white", alpha.f = .75), cex = .6)
```

# Monthly and annual timeseries, climatological normals

The functions `monthly()`, `annual()` nad `normal()` summarized the weather time series to monthly and annual time series and to climatological normal (long-term averages), respectively.
Summaries are calculated as follows:

  - $T_{min}$ is the minimum daily temperature recorded in the month or the year.
  - $T_{max}$ is the maximum daily temperature recorded in the month or the year.
  - $Prcp$ is the cumulative precipitation during the month or year.

`NA`s are removed during calculation.


```{r monthly}
monthly_ts <- monthly(daily_ts)
monthly_ts
plot(monthly_ts, "tmin")
```

```{r annual}
annual_ts <- annual(daily_ts)
annual_ts
plot(annual_ts, "tmin")
# normals <- normal(daily_ts) # to be implemented
```
