---
title: "Daily Weather Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Daily Weather Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GHCNr)
library(terra)  # for handling countries geometries
```

# Select GHCNd stations

The station inventory file of GHCNd is stored at <https://www.ncei.noaa.gov/products/land-based-station/global-historical-climatology-network-daily>.
The function `stations()` can read from this source or from a local file, specified with `filename`.
The inventory can also be downloaded to a file using `download_inventory()`.

```{r download-inventory, eval=FALSE}
inventory_file <- download_inventory("~/Downloads/ghcn-inventory.txt")
stations <- stations(inventory_file, variables = "TMAX")
stations
```

```{r download-inventory, echo=FALSE}
stations <- stations(variables = "TMAX")
stations
```
By specifying `variables = "TMAX"` only the stations that recorded that variable are kept.
Available variables implemented at the moment are precipitation ("PRCP"), minimum temperature ("TMIN"), and maximum temperature ("TMAX").

Stations spanning a time range can be filtered easily.
```{r filter-by-year}
stations <- stations[stations$startYear <= 1990, ]
stations <- stations[stations$endYear >= 2000, ]
stations
```

Spatial filters can also be easily applied.
Spatial boundaries of countries can be downloaded from <https://www.geoboundaries.org/> using the `get_countr(couuntry_code = ...)` function, where `country_code` is the ISO3 code.
```{r get-country}
italy <- get_country("ITA")
```
`get_countries()` can take several ISO3 codes to return a geometry of multiple countries.

```{r spatial-filter}
stations <- filter_stations(stations, italy)
plot(italy)
points(stations[, c("longitude", "latitude")], pch = 20, col = "dodgerblue")
```

# Download daily timeseries

Daily timeseries for a station can be downloaded using the `daily()` function.
In addition to the station ID, `daily()` needs start and end dates of the timeseries.
These should be provided as strings with the format "YYYY-mm-dd", e.g., "1990-01-01".

```{r daily}
daily_ts <- daily(
  stations$station[1],
  paste("1990", "01", "01", sep = "-"),  # shorten a bit as example
  paste(stations$endYear[1], "12", "21", sep = "-"),
  variables = "tmin"
)
daily_ts
```

Implmented variables are "tmin", "tmax", and "prcp".
`daily()` returns a table with the value of the variable chosen and associated flags.
Only certain flags from GHCNd are implemented, namely:
```{r flags, echo=FALSE}
GHCNr:::.flags()
```

Flagged records can be removed using `remove_flagged().
This will also remove the "*_flag=" column.
```{r remove-flagged}
daily_ts <- remove_flagged(daily_ts)
daily_ts
```

Coverage of the timeseries can be calculated using `daily_coverage()`.
```{r daily-coverage}
coverage <- daily_coverage(daily_ts)
```
The output is a table with coverage by month and year (`monthly_coverage`), by year (`annual_coverage`), and for the whole time period (`year`).
`annual_coverage` is constant within the same year and `year` is always a constant.

```{r coverages}
coverage[coverage$monthly_coverage < 1, ]
with(coverage[coverage$month == 1, ],
  barplot(
    annual_coverage,
    names.arg = year,
    las = 2
  )
)
```

Note that missing years are not shown and that `year_coverage` only calculates the number of years that have been covered, not their internal coverage.

```{r plot-timeseries, echo=FALSE, fig.width=6, fig.height = 4}
daily_ts$year <- as.numeric(format(daily_ts$date, "%Y"))
daily_ts$day <- as.numeric(format(daily_ts$date, "%j"))
palette <- hcl.colors(
  diff(range(unique(daily_ts$year))) + 1,
  "Zissou 1"
)
plot(
  c(1, 365), range(daily_ts$tmin, na.rm = TRUE),
  pch = NA, xlab = "Day of the year", ylab = "Tmin"
)
for (y in unique(daily_ts$year)) {
  with(
    daily_ts[daily_ts$year == y, ],
    lines(day, tmin, col = palette[as.numeric(y) - 1989], lw = 1.2)
  )
}
legend(
  x = 0, y = 20,
  legend = unique(daily_ts$year)[seq(1, 35, 5)],
  fill = palette[seq(1, 35, 5)]
)
```
